---
title: "rna-seq_opdracht3"
author: "Mathijs van Drongelen(1832719), Diana Tereshchenko(1865053)"
date: "`r Sys.Date()`"
output: html_document
---

```{r, setup}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE) 
```

```{r libraries, eval = TRUE, message=FALSE}
library(tidyverse)
library(Rsubread)
library(DESeq2)
library(pheatmap)
```
# Onderdeel Les 3 

```{r dds acquisition REMOVE FOR FINAL DOCUMENT}
Counting_table_raw<-readRDS("/home/daur2/rnaseq/rnaseq_ipsc/counts/read_counts.rds")

count_matrix <- Counting_table_raw$counts
metadata <- read_csv("/home/daur2/rnaseq/rnaseq_ipsc/ipsc_sampledata.csv")

# Convert the metadata to data frame object
metadata <- as.data.frame(metadata)

# Add rownames to the metadata data frame
rownames(metadata) <- paste0(metadata$Run, ".bam")

# Check if column names of count table are the same as row names of metadata object
colnames(count_matrix) == rownames(metadata)

#Factor Cell types as the difference in celltype / treatment to iPSC's is the point
# of interest

metadata$Cell_type<-metadata$Cell_type %>% str_replace("Skin derived fibroblast", "Skin_derived_fibroblast")

metadata$Cell_type <- metadata$Cell_type %>% 
factor(levels = c("Skin_derived_fibroblast", "iPSC"))


# Create the DESeqDataSet object
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = metadata, 
  design = ~ Cell_type
)
```

## Onderdeel 3a

- *Voer de DGE analyse uit met behulp van DESeq2. Hoeveel genen zijn er upgereguleerd (adjusted p-value < 0.05 en LFC > 1)? En hoeveel genen zijn er downgereguleerd (adjusted p-value < 0.05 en LFC < -1)?*

```{R, opdr3a, eval = TRUE}
# Perform the DGE analysis using DESeq2
ipsc_dge <- DESeq(dds)
# Obtain the results for the DGE analysis
ipsc_dge_results <- results(ipsc_dge, alpha = 0.05, lfcThreshold = 1)
summary(ipsc_dge_results)
```
- *Vergelijk de aantallen met de aantallen voor de airway studie. Wat is je conclusie?*
- Conclusie

## Onderdeel 3b
- *Create a Volcano plot in which you colour all genes with an adjusted p-value < 0.05 and |LFC|>1 darkblue. Also indicate the LFC and p-value thresholds as dashed lines in the plot.*

```{R, opdr3b, eval = TRUE}
# Create a data frame for plotting without genes with padj = NA
ipsc_dge_plotting <- data.frame(ipsc_dge_results) %>% filter(!is.na(padj))

# Create column specifying if gene is significantly differentially expressed
ipsc_dge_plotting <- ipsc_dge_plotting %>% 
  mutate(signif = if_else(padj < 0.05 & abs(log2FoldChange) > 1, 
                          "Significant", "Not significant"))
ipsc_dge_plotting

#Obtain the id of the most significantly upregulated gene
significant_genes <- ipsc_dge_results[which(ipsc_dge_results$padj < 0.05),]
topGene <- significant_genes[which.max(significant_genes$log2FoldChange),]
topGene_name <- rownames(topGene)
topGene_name

#Obtain the id of the most signigicantly downregulated gene
topGene_down <- significant_genes[which.min(significant_genes$log2FoldChange),]
topGene_down_name <- rownames(topGene_down)
topGene_down_name

#
# Create a volcano plot
ipsc_dge_plotting %>% 
  ggplot(aes(x = log2FoldChange, y = -log10(padj), color = signif)) +
  geom_point() + 
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") + 
  theme_minimal() +
  
  # Change the legend text and the colours
  scale_colour_manual(values = c("grey", "darkblue"), name = "Significance") +
  
  
  # Indicate thresholds as lines
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = -1, linetype = "dashed") +
  
    # Annotate the most significantly upregulated gene
  annotate("text", x = topGene$log2FoldChange, y = -log10(topGene$padj)*0.8, 
           label = topGene_name, colour = "blue", size = 2, vjust = -2) +
  annotate("text", x = topGene_down$log2FoldChange, y = -log10(topGene_down$padj)*0.8, 
           label = topGene_down_name, colour = "blue", size = 2, vjust = -2) 
```

## Onderdeel 3c
- *Maak een heatmap met daarin de count values voor de 15 significante DE genen (adjusted p-waarde < 0.05) met de grootste |LFC|.*'

```{R, opdr3c, eval = TRUE}

significant_genes <- ipsc_dge_results[which(ipsc_dge_results$padj < 0.05),]
top15_genes <- rownames(
  significant_genes[order(-abs(significant_genes$log2FoldChange)), ][1:15, ]
)

# Obtain the count values for these genes
count_values <- assay(dds)[top15_genes,]

# Replace the column names with conditions
colnames(count_values) <- colData(dds)$Cell_type

# Visualize the results in a heatmap
pheatmap(count_values, show_rownames = TRUE)
```


